---
title: "preliminary-analysis-3"
format: html
---

#### Set Up Document
```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
```

### The Question
How does land class effect species richness of a survey?

For this question, we are exploring the relationship between a quantitative variable and a categorical variable. Species richness is a discrete variable and land class is a nominal categorical variable.

For this analysis, we will be using one-way ANOVA to explore the relationship between our variables.

#### Read in the data
For this analysis we need to work with the whole gad_moon.csv dataset. We will need to use dplyr functions to create a species richness variable.
```{r}
gad <- read.csv(here("data", "gad_moon.csv"))

sp.richness.df <- gad |>
  group_by(LandClass, SpeciesName) |>
  summarise(individ.count = n()) |> 
  ungroup() |>
  group_by(LandClass) |>
  summarise(SpeciesRichness = n()) |>
  slice(1:11) |>
  mutate(LandClass = as.factor(LandClass))

sp.richness.df2 <- gad |>
  group_by(LandClass, SpeciesName, Date) |>
  summarise(n_individuals = n()) |> 
  ungroup() |>
  group_by(LandClass, Date) |>
  summarise(SpeciesRichness = n()) |>
  slice(1:279) |> 
  mutate(LandClass = as.factor(LandClass))

```
Now we have read in the amphibian data and created a species richness dataframe that includes the number of species found in each habitat. This code also changed LandClass to a factor data type and removed the NA land class variables.

The code for sp.richness.df2 creates observations per date per landclass, I'm not sure if I will use this but I will keep it for now.

#### plot the data
Make a barplot to explore the difference between the groups
```{r}
ggplot(sp.richness.df, aes(LandClass, SpeciesRichness))+
  geom_col()+
  coord_flip()
```
There is higher observed species richness in DevOpen, Woodwet, and EvergForest land classes. There is lower observed species richness in MixedForest, EmergHerbWet, and DevHigh land class.

#### build a model
```{r}
sp.richness.model <- lm(SpeciesRichness ~ LandClass, data = sp.richness.df2)
```
The model doesn't work with sp.richness.df so I used sp.richness.df2 instead. I think I will try to find a different approach to this question in the next stage of data analysis.

#### check assumptions
```{r}
autoplot(sp.richness.model)
```
This is a mess. A line is not the best way to model this data and it does not meet the assumption of normality.

#### Interpret the Model
```{r}
anova(sp.richness.model)
```
We can reject the null hypothesis that there is not an association between land class and species richness. Based on the ANOVA table, there is a significant relationship between LandClass and SpeciesRichness (F=3.4035, df=10, p=0.0003288).

```{r}
summary(sp.richness.model)
```
The summary table shows that there is a not difference between species richness in the deciduous forest land class and species richness in the other land classes.

We can use the multcomp package to better find out if any variables are different from each other  using a Tukey Test.
```{r}
library(multcomp)
phc1 <- glht(sp.richness.model, linfct = mcp(LandClass = "Tukey"))

cld(phc1)
```

The `cld()` fucntion shows that there are two distinct groups in the data that it labels a and b. Several of the land classes are not significantly distinct from either of the two groups and are labeled ab.

DevLow and DevMed are in group b, while DevOpen is in group a. The other groups are in group ab.

#### plot again
I want to get the mean species richness/date to add to the plot
```{r}
sr_means <- sp.richness.df2 |>
  group_by(LandClass) |>
  summarise(mean.sr = mean(SpeciesRichness)) 

ggplot(data = sp.richness.df2, aes(x = LandClass, y = SpeciesRichness))+
  geom_point(aes(color = LandClass))+
  geom_point(data = sr_means, aes(x = LandClass, y = mean.sr,
                                  fill = LandClass),
             size = 5, shape = 23)+
  coord_flip()+
  theme_bw()
```

#### A note
I know that this analysis is a mess and likely needs redone. I expect to come back to this after break and fix this, however, I wanted to be able to submit this assignment before I started travelling so I'd have one fewer thing to worry about.