---
title: "preliminary-analysis-1"
format: html
---

#### Set Up Document

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
```

### The Question

How does moon illumination effect the number of amphibians observed?

In this question, we are exploring the relationship between two integers. The number of amphibians found is discrete data and and the percent moon illumination is continuous ratio data.

When exploring the relationship between these data types, the best approach to use is a simple linear regression model.

#### Read in data

```{r}
gad_totals <- read.csv(here("data", "gad_totals1.csv"))
glimpse(gad_totals)
```

After viewing the dataset, you can see that several data types were read in as the wrong type. The `write.csv()` fucntion also created the useless X variable that can be removed. Date should be a dtae variable and phase should be a factor. These variables won't be used for this analysis but I will change them anyways.

```{r}
gad_totals <- gad_totals |>
  select(-X) |>
  mutate(Date = ymd(Date)) |>
  mutate(phase = as.factor(phase))
```


#### plot the data

I will start by making a simple scatterplot looking at the relationship between moon illumination and number of amphibians found.

```{r}
ggplot(gad_totals, aes(x = illum_pct, y = n_found))+
  geom_point()+
  theme_bw()
```

Based on this scatterplot,there doesn't seem to be a strong relationship between the number of amphibians found and moon illumination, however, this analysis doesn't yet account for survey effort.

The y-intercept of the regression line is likely around 10.

The slop of the line is likely a little higher than 0, maybe around 0.1.

#### Build a model

```{r}
illum_model <- lm(n_found ~ illum_pct, data = gad_totals)
```

This model has an intercept of 9.93664 which is close to my prediction and a slope of 0.02186 which is lower than I expected.

#### Check assumptions

```{r}
autoplot(illum_model)
```

The Residuals vs Fitted, Scale-Location, and Residuals vs Leverage plots all look good. The Normality plot, however, is not good and does not meet the assumption of normality.

For sake of this project, I will continue the analaysis. Maybe a transformation can be applies to the data at a later stage of the project to fix this.

#### Interpret results

```{r}
anova(illum_model)
```

Based on these results, we fail to reject the null hypothesis that there is no relationship between moon phase and number of amphibians found while surveying (F=0.3304, df=1, p=0.5662).

Because there is no relaionship between the variables, the analysis endes here and I will not use the `summary()` function to look at significance of predictors.

#### Plot Again

```{r}
ggplot(gad_totals, aes(illum_pct, n_found))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x = "Moon Illumination Percent",
       y = "Number of Amphibians Found")+
  theme_bw()
```
