---
title: "CAPDAP cleanup"
format: html
---

### set up
```{r}
rm(list = ls())
library(tidyverse)
library(here)
```

### read in datasets
```{r}
gad <- read.csv(here("data", "gad_cleaned.csv"))
moon <- read.csv(here("data", "moon_illumination_1800-2100.csv"))
```

need to add the moon phase data to the amphibian data, use dplyr joins to do so. First, the date categories must be in the same format to use as a key.
```{r}
glimpse(moon)
glimpse(gad)
```

the date variable for moon has both date and time in the same variable. They need to be seperated before you can use lubridate functions.

```{r}
moon <- moon |> 
  separate(date, c("date", "time"), sep = " ") |>
  select(-time, -X)
```

now both date variables can be converted to dates but they must be in the same format
```{r}
gad$Date <- mdy(gad$Date)
moon$date <- ymd(moon$date)
```


join the dates using dplyr functions
```{r}
gad_moon <- left_join(gad, moon, join_by(Date == date))
```

export the new dataset as a csv file
```{r}
write.csv(gad_moon, 
          file = here("data/gad_moon.csv"), 
          row.names = F)
```

