---
title: "preliminary-analysis-2"
format: html
---

#### Set Up Document
```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
```

### The Question

How does moon phase effect the number of amphibians observed?

In this question we are exploring the relationship between a categorical variable and a quantitative variable. Moon phase is a nominal categorical variable and number of amphibians observed is a discrete data type.

When exploring relationships between one categorical predictor and a quantititive response, you can use a one-way ANOVA test.

#### Read in data
I need to use my gad_totals dataset for this analysis.
```{r}
gad_totals <- read.csv(here("data", "gad_totals1.csv"))
glimpse(gad_totals)
```
Like in the first analysis, the dataset needs cleaned up. This time it is important that the phase variable is a factor rather than a character data type. I will also remove X and fix the time variable. When looking at the full dataset, there is a row at the end with data only for count, I'll also remove that point.
```{r}
gad_totals <- gad_totals |>
  select(-X) |>
  mutate(Date = ymd(Date)) |>
  mutate(phase = as.factor(phase)) |>
  slice(1:154)
```

#### plot the data
I'll start by making a boxplot to view the relationship between moon phase and amphibians found and estimate the group means.
```{r}
ggplot(gad_totals, aes(x = phase, y = n_found))+
  geom_boxplot()+
  geom_point()+
  coord_flip()+
  theme_bw()
```
Based on this boxplot, the means are very close together. All of the groups seem to have means between 0 and 10. I won't give exact estimates because they seem to be so close together that they will all be relatively similar. I don't expect there to be a difference in means between the 8 groups.

#### construct the model
use the lm function to build a model for anova
```{r}
phase_model <- lm(n_found ~ phase, data = gad_totals)
```

#### check assumptions
```{r}
autoplot(phase_model)
```
The assumptions do not look great, I'll move on anyways.

#### Interpret model
```{r}
anova(phase_model)
```
Based on the ANOVA table, we fail to reject the null hypothesis that there is no relationship between moon phase and number of amphibians found (F = 1.1909, DF = 7, 
P = 0.3115).

#### Plot again

To make an improved plot, I need a dataset that just includes the mean for each group.
```{r}
phase_means <- gad_totals |>
  group_by(phase) |>
  summarise(mean_found = mean(n_found))
```

Now I can plot the data.
```{r}
ggplot(data = gad_totals, aes(x = phase, y = n_found, color = phase))+
  geom_point(alpha = 0.75, size = 3)+
  geom_point(data = phase_means, aes(x=phase, y=mean_found, fill=phase),
             size = 5, shape = 23)+
  coord_flip()+
  labs(x = "Moon Phase",
       y = "Number of Amphibians Found")+
  theme_bw()
  
```

