---
title: "final-analysis-3"
format: html
---

#### Set Up Document
```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
```

### The Question
How does land class effect species richness of a survey?

For this question, we are exploring the relationship between a quantitative variable and a categorical variable. Species richness is a discrete variable and land class is a nominal categorical variable.

For this analysis, we will be using one-way ANOVA to explore the relationship between our variables.

#### Read in data
For this analysis we need to work with the whole gad_moon.csv dataset. We will need to use dplyr functions to create a species richness variable.
```{r}
gad <- read.csv(here("data", "gad_moon.csv"))

sp.richness.df2 <- gad |>
  group_by(LandClass, SpeciesName, Date) |>
  summarise(n_individuals = n()) |> 
  ungroup() |>
  group_by(LandClass, Date) |>
  summarise(SpeciesRichness = n()) |>
  na.omit() |> 
  mutate(LandClass = as.factor(LandClass))
```

The code for sp.richness.df2 creates observations per date per landclass. It also removes all na values and puts the variables into the correct data types.

#### plot the data

Make a boxplot of the means and standard deviations of each group.
```{r}
ggplot(sp.richness.df2, aes(LandClass, SpeciesRichness))+
  geom_boxplot()+
  coord_flip()

#group the ones with one variable into other
```
There is a clear difference in means between the groups. some of the groups have far fewer observations than others. I will combine the groupings that have too few observations and reasses. DevOpen will be combine with DevMed, DevLow, and DevHigh to form a new category called Developed. OpenWater, Mixed Forest, and EmergHerbWet will go into an "Other" category. 
```{r}
sp.richness.df2.grouped <- sp.richness.df2 |> 
  mutate(LandClass = fct_collapse(LandClass, Developed = c("DevOpen",
                                                           "DevMed",
                                                           "DevLow",
                                                           "DevHigh"))) |>
  mutate(LandClass = fct_collapse(LandClass, Other = c("OpenWater",
                                                   "MixedForest",
                                                   "EmergHerbWet")))
  
```

Now I can replot
```{r}
ggplot(sp.richness.df2.grouped, aes(LandClass, SpeciesRichness))+
  geom_boxplot()+
  coord_flip()
```
The developed category looks good, the other category still isn't great but I'll work with it. Now I can estimate the group mean species richness/survey.
WoodWet - 2.1
PastHay - 0.8
EvergForest - 1.4
Other - 0.8
Developed - 0.8
DecidForest - 2.1

I also want to look at the distribution of the data.
```{r}
ggplot(sp.richness.df2.grouped, aes(SpeciesRichness))+
  geom_histogram()+
  facet_wrap(~LandClass,
             ncol = 1)+
  theme()
```
The data is not normally distributed, so a GLM should be used rather than a general linear model.

#### build a model
```{r}
sr.glm <- glm(SpeciesRichness ~ LandClass, data = sp.richness.df2.grouped, family = poisson)
```

#### check assumptions
```{r}
autoplot(sr.glm)
```
The plot doesn't meet the assumptions for a regular linear model so using a glm was correct. The normality assumption is very shaky, likely due to zero inflation in the data.

#### evaluate the model
```{r}
anova(sr.glm, test = "Chisq")

summary(sr.glm)
```

The total deviance of the data is 174.53, while the amount of deviance in the land class is 168.78 This means that 3.2% of deviance in the species richness can be explained by land class. The p-value is also very high (p=0.3314) so we fail to reject the null hypothesis that there is no association between DEC land class and species richness per survey.


#### plot again
I want to get the mean species richness/date to add to the plot
```{r}
sr_means <- sp.richness.df2.grouped |>
  group_by(LandClass) |>
  summarise(mean.sr = mean(SpeciesRichness)) 

ggplot(data = sp.richness.df2.grouped, aes(x = LandClass, y = SpeciesRichness))+
  geom_point(aes(color = LandClass))+
  geom_point(data = sr_means, aes(x = LandClass, y = mean.sr,
                                  fill = LandClass),
             size = 5, shape = 23)+
  coord_flip()+
  theme_bw()
```